---
title: "Categorical Sequence Analysis with Optimal Matching: An Application with Data from the 'Women 40+ Healthy Aging Study'"
author: 'Adriana Clavijo Daza'
date: 'June 2023'
institution: 'University of Bern'
division: 'Faculty of Science'
advisor: 'Prof. Dr. David Ginsbourger'
altadvisor: 'Dr. Serena Lozza-Fiacco'
department: 'Institute of Mathematical Statistics and Actuarial Science'
degree: 'Master of Science in Statistics and Data Science'
knit: bookdown::render_book
# site: bookdown::bookdown_site

# The next two lines allow you to change the spacing in your thesis. You can 
# switch out \onehalfspacing with \singlespacing or \doublespacing, if desired.
header-includes:
    - \usepackage{setspace}\onehalfspacing
    - \usepackage{algpseudocode}
    - \usepackage{algorithm}
    - \AtBeginDocument{\renewcommand{\chaptername}{Section}}

# Remove the hashtag to specify which version of output you would like.
# Can only choose one at a time.
output:
  thesisdown::thesis_pdf: default 
#  thesisdown::thesis_gitbook: default         
#  thesisdown::thesis_word: default
#  thesisdown::thesis_epub: default

# If you are creating a PDF you'll need to write your preliminary content 
# (e.g., abstract, acknowledgements) below or use code similar to line 25-26 
# for the .RMD files. If you are NOT producing a PDF, delete or silence
# lines 25-39 in this YAML header.
abstract: |
  `r if(knitr:::is_latex_output()) paste(readLines("prelims/00-abstract.Rmd"), collapse = "\n  ")`

# If you'd rather include the preliminary content in files instead of inline
# like below, use a command like that for the abstract above.  Note that a tab 
# is needed on the line after the `|`.
acknowledgements: |
  I want to thank a few people.
dedication: |
  You can have a dedication here if you wish. 

# Note that abbreviations in lowercase letters will NOT be automatically
# capitalized.
# abbreviations:
#   ABC: American Broadcasting Company
#   CBS: Colombia Broadcasting System
#   CUS: Computer User Services
#   PBS: Public Broadcasting System
#   NBC: National Broadcasting Company

# Specify the location of the bibliography below
bibliography: bib/thesis.bib

# Download your specific csl file and refer to it in the line below.
csl: csl/apa.csl
lot: true
lof: true
---

<!--
Above is the YAML (YAML Ain't Markup Language) header that includes a lot of 
metadata used to produce the document.  Be careful with spacing in this header!

If you'd prefer to not include a Dedication, for example, simply delete the 
section entirely, or silence them (add # before each line). 

If you have other LaTeX packages you would like to include, delete the # before
header-includes and list the packages after hyphens on new lines.

If you'd like to include a comment that won't be produced in your resulting file
enclose it in a block like this.

If you receive a duplicate label error after knitting, make sure to delete the
index.Rmd file and then knit again.
-->

```{r packages, include = FALSE}
library(readxl)
library(tidyr)
library(dplyr)
library(stringr)
library(ggplot2)
library(TraMineR)
library(TraMineRextras)
library(cluster)
library(extrafont)
```

```{r helpers, include = FALSE}
## Note: this is order dependent.
source("helpers/distance_matrix.R", local = knitr::knit_global())
source("helpers/predict_scores.R", local = knitr::knit_global())
source("helpers/get_MSE.R", local = knitr::knit_global())
```

```{r setup, include = FALSE}
# Global knitr chunk options
## Note: updating global chunk options propagates, invalidating existing cache.
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  # Graphics settings
  fig.align = "center",
  fig.width = 7.5,
  fig.height = 5.25,
  ## Use cairo_pdf to embed fonts in the plots (and preserve it when rendering):
  ## https://stackoverflow.com/a/36663975
  dev = "cairo_pdf",
  ## Font family for base plots. Requires it available via extrafont::
  dev.args = list(family = "Latin Modern Math") ## For base plots.
)

extrafont::loadfonts(device = "all")
# GGplot theme.
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 12))

ggplot2::theme_update(
  ## Default axis labels size is 80% of plot font size. Increasing to 100%.
  axis.text = ggplot2::element_text(size = ggplot2::rel(1)),
  ## Font family. Requires it available via extrafont::
  text = ggplot2::element_text(family = "Latin Modern Math")
)

options(width = 80)
```

<!--
The acknowledgments, preface, dedication, and abstract are added into the PDF
version automatically by inputting them in the YAML at the top of this file.
Alternatively, you can put that content in files like 00--prelim.Rmd and
00-abstract.Rmd like done below.
-->

```{r eval = !knitr::is_latex_output(), child = here::here("prelims", "00--prelim.Rmd")}
```

```{r eval = !knitr::is_latex_output(), child = here::here("prelims", "00-abstract.Rmd")}
```

```{r, child = "data_prep.Rmd"}
```

```{r, child = "main_analysis.Rmd"}
```

<!-- The {.unnumbered} option here means that the introduction will be 
"Chapter 0." You can also use {-} for no numbers on chapters.
-->

# Introduction {.unnumbered}

Since optimal matching was introduced by @Abbott-1986, it has been extensively used in social studies to answer questions involving processes that take values in a categorical set of states and occur along a specific period of time. For instance, in sociology, optimal matching has been employed in several studies tracking the professional development of specific groups of people, see @Chan-1995, @Torsten-2014 or @Gubler-2015, and to analyze life course data, for example, @Widmer-2009 or @Bastin-2015.

The common result of several studies employing optimal matching is to find similarities and dissimilarities among categorical sequences and identify groups of trajectories that exhibit similar patterns. This is achieved through the application of clustering techniques (usually hierarchical) to the distance matrix generated with help of the algorithm. However, critics as @Wu-2000 point out that optimal matching has limitations. A notable flaw lies in the definition of the cost matrix and the strong assumption of its symmetry.

In this work, we explore the potential to obtain predictions of additional variables, namely personality scores, based on the information obtained from the sequential data via optimal matching. For this purpose, we use a dataset that contains categorical sequences of the relationship history of a group of women in the context of the 'Women 40+ Healthy Aging Study' and use the implementation of the algorithm in the `R` package `TraMineR` (@TraMineR). 

<!-- TODO: use section labels: 
       bookdown.org/yihui/bookdown/cross-references.html -->
In [Section 1](#om), we provide a detailed description of the optimal matching algorithm, the methods considered to obtain the cost matrix and the normalization methods for the distance matrix. In [Section 2](#data), we introduce the data used in this work and perform optimal matching to obtain clusters of sequences and use visualizations to provide a description of the groups. In [Section 3](#knn), we describe $k$-nearest neighbors ($k$-NN), the method we chose for prediction of the personality scores based on the data from the previous section. In [Section 4](#scenarios), we show some of the additional scenarios considered, specifically those that produce the best predictions for each personality trait. Finally, we present some conclusions and recommendations for future work.
